### 2023-07-07 docs/intro.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  intro.tex
%
%  Contains introduction and quick start for users.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2016-2019 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Quick Start for Users}

\subsection{Running an Example}

\subsubsection{Input file}

The behaviour of \textit{Ludwig} at run time is controlled by
a plain text input file which consists of a series for key value
pairs. Each key value pair controls a parameter or parameters
within the code, for example the system size and number of time steps,
the fluid properties density and viscosity, the free energy type
and associated parameters (if required), frequency and type of
output, parallel decomposition, and so on.

For most keys, the associated property has some default value which
will be used by the code if the relevant key is not present (or
commented out) in the input file. While such default values are
chosen to be at least sensible, users need to be aware that all
necessary keys need to be considered for a given problem. However,
many keys are irrelevant for any given problem.

A significant number of example input files are available as
part of the test suite, and these can form a useful starting
point for a given type of problem. We will consider one which
uses some of the common keys.

\subsubsection{Example input}

We will consider a simple example which computes the motion of
a single spherical colloidal particle in an initially stationary
fluid of given viscosity. The velocity may be measured as a
function of time. Assume we have an executable in the \texttt{src}
directory.

Make a copy of the input file
\begin{lstlisting}
$ cp ../tests/regression/d3q19-extra/serial-auto-c01.inp .
\end{lstlisting}

Inspection of this file will reveal the following: blank lines
and comments --- lines beginning with \texttt{\#} --- may be included,
but are ignored at run time; other lines are parsed as key value
pairs, each pair on a separate line, and with key and value separated
by one or more spaces. Values may be characters or strings, a
scalar integer or 3-vector of integers, or a scalar or 3-vector
of floating point numbers. Values which are 3-vectors are delineated
by an underscore character (not a space), e.g., \texttt{1\_2\_3},
and always correspond to a Cartesian triple $(x,y,z)$.
A given value is parsed appropriately in the context of the associated key.

So, the first few lines of the above file are:
\lstinputlisting[firstline=1, lastline= 16]
{../tests/regression/d3q19-extra/serial-auto-c01.inp}
Here, the first key value pair \texttt{N\_cycles 40} sets the number
of time steps to 40, while the second, \texttt{size 64\_64\_64}, sets
the system size to be 64 points in each of the three coordinate
directions. The \texttt{grid} key relates to the parallel
domain decomposition as is discussed in the following
section.

\subsubsection{Run time}

The executable takes a single argument on the command line which
is the name of the input file, which should be in the same
directory as the executable. If no input file name is supplied,
the executable will look for a default \texttt{input}. If no
input file is located at all, an error to that effect will be
reported.

\textbf{Serial: }
If a serial executable is run in the normal way with the copy of the
input file as the argument the code should take around 10--20 seconds
to execute 40 steps. The fist few lines of output should look like:

\begin{lstlisting}[belowskip=0pt]
$ ./Ludwig.exe ./serial-auto-c01.inp
\end{lstlisting}
\lstinputlisting[aboveskip=0pt, firstline=1, lastline= 16]
{../tests/regression/d3q19-extra/serial-auto-c01.log}
This output shows that the appropriate input file has been read, and
the system size set correspondingly with a number of default settings
including periodic boundary conditions. ``No free energy'' tells us we
are using a single Newtonian fluid.

Normal termination of execution is accompanied by a report
of the time take by various parts of the code, and finally by
\begin{lstlisting}
...
Ludwig finished normally.
\end{lstlisting}

\textbf{Parallel: }
The executable compiled and linked against the MPI library can
be run with the MPI launcher for the local system, often
\texttt{mpirun}. For example, a run on 8 MPI tasks produces a
similar output to that seen in the serial case, but reports
details of the local domain decomposition:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe ./serial-auto-c01.inp
Welcome to Ludwig v0.1.26 (MPI version running on 8 processes)
...
System details
--------------
System size:    64 64 64
Decomposition:  2 2 2
Local domain:   32 32 32
...
\end{lstlisting}
The decomposition is controlled by the \texttt{grid} key in
the input. Here, \texttt{grid 2\_2\_2} is consistent with the
8 MPI tasks specified on the command line, and the resulting
local decomposition is 32 lattice points in each coordinate direction.
If no grid is specified in the input, or a request is
make for a grid which cannot be met (e.g., the product
of the grid dimensions does not agree with the total number
of MPI tasks available)
the code will try to determine its own decomposition as best
it can. If no valid parallel decomposition is available at all,
the code will exit with a message to that effect.

Further details of various input key value pairs are given in
relevant sections of the documentation.

\subsubsection{Output}

The standard output of the running code produces a number of
aggregate quantities which allow a broad overview of the progress
of the computation to be seen by the user. These include, where
relevant, global statistics related to fluid density and momentum,
the integrated free energy, particle-related quantities, and so on.
The frequency of this information can be adjusted from the input
file (see, e.g., \texttt{freq\_statistics}).

Output for lattice-based quantities (for example, the velocity field
$u(\mathbf{r}; t)$) is via external file. This output is in serial
or in parallel according to how the model is run, and may be in
either ASCII or raw binary format as required. Output files are produced
with time step encoded in the file, and an extension which describes the
parallel output decomposition. Further, each quantity output in this way
is accompanied by a metadata description with \texttt{meta} extension.
For example, the two output files
\begin{lstlisting}
bash-3.2$ ls dist*
dist-00000020.001-001   dist.001-001.meta
\end{lstlisting}
contain the LB distributions for time step 20 (the file is 1 of a total
number of 1 in parallel), and the plain text metadata description,
respectively.

To ensure output for based-based quantities is in the correct order for
analysis, post-processing
may be required (always if the code is run in parallel). This uses a utility
provided for the purpose which required both the data and the metadata
description to recombine the parallel output. This utility is described
ELSEWHERE.

Output for colloidal particle data is again to file with a name encoding
the time step and parallel decomposition of the output. For example,
\begin{lstlisting}
bash-3.2$ ls config*
config.cds00000020.001-001
\end{lstlisting}
contains particle data for time step 20 in a format described in
Section~\ref{section-examples}.
These data may be requested in ASCII or raw binary format.

\subsubsection{Errors and run time failures}

The code attempts to provide meaningful diagnostic error messages
for common problems. Such errors include missing or incorrectly
formatted input files, inconsistent input values, and unacceptable
feature combinations. The code should also detect other run time
problems such as insufficient memory and errors writing output
files. These errors will result in termination.

Instability in the computation will often be manifested by numerically
absurd values in the statistical output (ultimately \texttt{NaN} in
many cases). Instability may or may not result in termination, depending
on the problem. Such instability is very often related to poor parameter
choices, of which there can be many combinations. Check the
relevant section of the documentation for advice on reasonable starting
parameters for different problems.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  intro.tex
%
%  Contains introduction and quick start for users.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2016-2023 The University of Edinburgh
%
%  This material is being migrated to https://ludwig.epcc.ed.ac.uk/index.html
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@
